The Sound of Teaching Music: Expert Pianists' Performance Modulations for Novices

### 2021-06-17 paper/manuscript.Rmd
updated methods (exp2)
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

We recruited 36 piano experts who played the piano for at least the past 10 years or were studying advanced piano performance at a music school at the time of recruitment. In order to ensure that participants had sufficient motor skills to perform the piece, we asked them to practise the piece with the non-expressive sheet music (*Figure \ref{fig:stimuli}, A*) in the lab. Once they could perform the piece without pitch errors twice consecutively within 5 attempts, the experiment started. Those who could not meet this requirement were excluded from data analysis (as a result, none of the participants were excluded based on this criterion). For data analysis, we excluded three participants due to experimental errors, and two participants because they deviated substantially from the prescribed tempo (outside 2 standard deviations from the average tempo across participants). Thirty-one participants (`r nrow(dt_included[Gender == "Female"])` female) were included in data analysis. Most participants were right-handed (left; `r nrow(dt_included[Handedness == "Left"])`, ambidextrous; `r nrow(dt_included[Handedness == "Both"])`). They had `r round(mean(dt_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_included$PianoTotalPractice), 2)`). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

We recruited 21 piano experts who already had a degree in (above bachelor or equivalent) in piano performance/teaching or were studying piano performance at a music school at the time of recruitment. For data analysis, we excluded 1 participants due to insufficient motor skills (1) and an overall deviated tempo (1; outside 2 standard deviations from the average tempo across participants). Nineteen participants (`r nrow(dt_2_included[Gender == "Female"])` female) were included in data analysis. Most participants were right-handed (left; `r nrow(dt_2_included[Handedness == "Left"])`). They had `r round(mean(dt_2_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_2_included$PianoTotalPractice), 2)`).

@@

The same apparatus as Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contained our targeted expressions (i.e., articulation, dynamics) and was relatively simple in terms of motor skills. The first 12 measures of the original piece was used and modified so that it had the almost equal number of data points for each dependent variable. The stimulus consisted of a 12-measure isochronous melody notated in 4/4 meter. Non-expressive sheet was used for the purpose of practice (*Figure \ref{fig:stimuli-2}, A*). Expressive notations were added to the non-expressive sheet music for the experiment (*Figure \ref{fig:stimuli-2}, B-C*). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., *Figure \ref{fig:stimuli-2}, A*) prior to the experiment. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations (see details in *[Supplemental Materials](#supplemental)*). Third, participants could choose their preferred tempo from one of the 3 options (100, 110 and 120 beats per minute) as a leading metronome. In order to make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the non-expressive sheet music. As Experiment 1, those who could not perform the piece twice consecutively within 5 attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

Data cleaning, preprocessing and statistical analysis were almost identical as Experiment 1. For statistical analysis, only 8th notes with expressive notations were included. As a result, only one 8th note in the 4th measure without any expression was not included to calculate dependent variables. IOIs were normalised by their preferred tempo because each participant chose different tempi from the 3 options. Key-overlap Ratio (KOR) was calculated by dividing KOT by the mean IOI of each performance in order to normalise KOT due to a different tempo for each participant. Additionally, we include KV difference (i.e., KV difference for each interval) at transition points (i.e., forte to piano or piano to forte) as a dependent variable based on the findings of Experiment 1. Three trials were entirely excluded from data analysis because participants did not follow sheet music. With the same method as Experiment 1, pitch errors were removed manually. For onsets, `r round(sum(error_onset_summary_2$N)/((31*4*18)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes; `r round(error_onset_summary_2[Reason == "Extra Notes"]$N/((31*4*18)-3)*100, 2)`%, missing notes; `r round(error_onset_summary_2[Reason == "Missing Notes"]$N/((31*4*18)-3)*100, 2)`%, substituted notes; `r round(error_onset_summary_2[Reason == "Substituted"]$N/((31*4*18)-3)*100, 2)`%). For offsets, `r round(sum(error_offset_summary_2$N)/((31*4*18)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes; `r round(error_offset_summary_2[Reason == "Extra Notes"]$N/((31*4*18)-3)*100, 2)`%, missing notes; `r round(error_offset_summary_2[Reason == "Missing Notes"]$N/((31*4*18)-3)*100, 2)`%, substituted notes; `r round(error_offset_summary_2[Reason == "Substituted"]$N/((31*4*18)-3)*100, 2)`%). Removed outliers (i.e., responses outside 3 standard deviations from the mean) were less than 5% of overall responses for each dependent variable.

>>>>

We recruited 36 piano experts who played the piano for at least the past 10 years or were studying advanced piano performance at a music school at the time of recruitment. In order to ensure that participants had sufficient motor skills to perform the piece, we asked them to practise the piece with the non-expressive sheet music (*Figure \ref{fig:stimuli}, A*) in the lab. Once they could perform the piece without pitch errors twice consecutively within 5 attempts, the experiment started. Those who could not meet this requirement were excluded from data analysis (as a result, none of the participants were excluded based on this criterion). For data analysis, we excluded three participants due to experimental errors, and two participants because they deviated substantially from the prescribed tempo (outside 2 standard deviations from the average tempo across participants). Thirty-one participants (`r nrow(dt_included[Gender == "Female"])` female) were included in data analysis. Most participants were right-handed (left: `r nrow(dt_included[Handedness == "Left"])`, ambidextrous: `r nrow(dt_included[Handedness == "Both"])`). They had `r round(mean(dt_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_included$PianoTotalPractice), 2)`). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

We recruited 21 piano experts who already had a degree in (above bachelor or equivalent) in piano performance/teaching or were studying advanced piano performance at a music school at the time of recruitment. For data analysis, we excluded one participants due to insufficient motor skills, and one because of large tempo deviations (outside 2 standard deviations from the average tempo across participants). Nineteen participants (`r nrow(dt_2_included[Gender == "Female"])` female) were included in the data analysis. Most participants were right-handed (left: `r nrow(dt_2_included[Handedness == "Left"])`). They had `r round(mean(dt_2_included$PianoTotalPractice), 2)` years of practice on average (*SD* = `r round(sd(dt_2_included$PianoTotalPractice), 2)`).

@@

The same apparatus as in Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contained our targeted expressions (i.e., articulation, dynamics) and was relatively simple in terms of motor skills. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in 4/4 meter. Non-expressive sheet music was used for the purpose of practice (*Figure \ref{fig:stimuli-2}, A*). Expressive notations were added to the non-expressive sheet music for the experiment (*Figure \ref{fig:stimuli-2}, B-C*). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as in Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., *Figure \ref{fig:stimuli-2}, A*) prior to the experiment. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations (see details in *[Supplementary Materials](#supplemental)*). Third, participants could choose their preferred tempo from one of three options (100, 110 and 120 beats per minute). The chosen tempo was again cued by a leading metronome. In order to make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the non-expressive sheet music. As Experiment 1, those who could not perform the piece twice consecutively within 5 attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

Data cleaning, preprocessing and statistical analysis were almost identical to Experiment 1. For statistical analysis, only 8th notes with expressive notations were included. As a result, only one 8th note in the 4th measure without any expression was not included. IOIs were normalised by their preferred tempo because each participant chose a tempo from the three options. Given the different tempi, a key-overlap ratio (KOR) was calculated by dividing KOT by the mean IOI of each performance in order to normalise KOT. Additionally, we include KV difference (i.e., KV difference for each interval) at transition points (i.e., forte to piano or piano to forte) as a dependent variable based on the findings of Experiment 1. Three trials were entirely excluded from data analysis because participants did not follow the sheet music. Using the same approach as in Experiment 1, pitch errors were removed manually. For onsets, `r round(sum(error_onset_summary_2$N)/((31*4*18)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes; `r round(error_onset_summary_2[Reason == "Extra Notes"]$N/((31*4*18)-3)*100, 2)`%, missing notes; `r round(error_onset_summary_2[Reason == "Missing Notes"]$N/((31*4*18)-3)*100, 2)`%, substituted notes; `r round(error_onset_summary_2[Reason == "Substituted"]$N/((31*4*18)-3)*100, 2)`%). For offsets, `r round(sum(error_offset_summary_2$N)/((31*4*18)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes; `r round(error_offset_summary_2[Reason == "Extra Notes"]$N/((31*4*18)-3)*100, 2)`%, missing notes; `r round(error_offset_summary_2[Reason == "Missing Notes"]$N/((31*4*18)-3)*100, 2)`%, substituted notes; `r round(error_offset_summary_2[Reason == "Substituted"]$N/((31*4*18)-3)*100, 2)`%). For each dependent variable, removing outliers (i.e., responses outside 3 standard deviations from the mean) resulted in less than 5% of overall responses being removed.

### 2021-06-17 paper/manuscript.tex
updated methods (exp2)
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

We recruited 36 piano experts who played the piano for at least the past 10 years or were studying advanced piano performance at a music school at the time of recruitment. In order to ensure that participants had sufficient motor skills to perform the piece, we asked them to practise the piece with the non-expressive sheet music (\emph{Figure \ref{fig:stimuli}, A}) in the lab. Once they could perform the piece without pitch errors twice consecutively within 5 attempts, the experiment started. Those who could not meet this requirement were excluded from data analysis (as a result, none of the participants were excluded based on this criterion). For data analysis, we excluded three participants due to experimental errors, and two participants because they deviated substantially from the prescribed tempo (outside 2 standard deviations from the average tempo across participants). Thirty-one participants (15 female) were included in data analysis. Most participants were right-handed (left; 2, ambidextrous; 2). They had 12.45 years of practice on average (\emph{SD} = 5.66). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

We recruited 21 piano experts who already had a degree in (above bachelor or equivalent) in piano performance/teaching or were studying piano performance at a music school at the time of recruitment. For data analysis, we excluded 1 participants due to insufficient motor skills (1) and an overall deviated tempo (1; outside 2 standard deviations from the average tempo across participants). Nineteen participants (9 female) were included in data analysis. Most participants were right-handed (left; 2). They had 15.65 years of practice on average (\emph{SD} = 5.67).

@@

The same apparatus as Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contained our targeted expressions (i.e., articulation, dynamics) and was relatively simple in terms of motor skills. The first 12 measures of the original piece was used and modified so that it had the almost equal number of data points for each dependent variable. The stimulus consisted of a 12-measure isochronous melody notated in 4/4 meter. Non-expressive sheet was used for the purpose of practice (\emph{Figure \ref{fig:stimuli-2}, A}). Expressive notations were added to the non-expressive sheet music for the experiment (\emph{Figure \ref{fig:stimuli-2}, B-C}). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., \emph{Figure \ref{fig:stimuli-2}, A}) prior to the experiment. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations (see details in \emph{\protect\hyperlink{supplemental}{Supplemental Materials}}). Third, participants could choose their preferred tempo from one of the 3 options (100, 110 and 120 beats per minute) as a leading metronome. In order to make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the non-expressive sheet music. As Experiment 1, those who could not perform the piece twice consecutively within 5 attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

Data cleaning, preprocessing and statistical analysis were almost identical as Experiment 1. For statistical analysis, only 8th notes with expressive notations were included. As a result, only one 8th note in the 4th measure without any expression was not included to calculate dependent variables. IOIs were normalised by their preferred tempo because each participant chose different tempi from the 3 options. Key-overlap Ratio (KOR) was calculated by dividing KOT by the mean IOI of each performance in order to normalise KOT due to a different tempo for each participant. Additionally, we include KV difference (i.e., KV difference for each interval) at transition points (i.e., forte to piano or piano to forte) as a dependent variable based on the findings of Experiment 1. Three trials were entirely excluded from data analysis because participants did not follow sheet music. With the same method as Experiment 1, pitch errors were removed manually. For onsets, 3.32\% of the trials contained at least one pitch error (extra notes; 1.66\%, missing notes; 1.57\%, substituted notes; 0.09\%). For offsets, 5.11\% of the trials contained at least one pitch error (extra notes; 1.66\%, missing notes; 1.57\%, substituted notes; 1.88\%). Removed outliers (i.e., responses outside 3 standard deviations from the mean) were less than 5\% of overall responses for each dependent variable.

>>>>

We recruited 36 piano experts who played the piano for at least the past 10 years or were studying advanced piano performance at a music school at the time of recruitment. In order to ensure that participants had sufficient motor skills to perform the piece, we asked them to practise the piece with the non-expressive sheet music (\emph{Figure \ref{fig:stimuli}, A}) in the lab. Once they could perform the piece without pitch errors twice consecutively within 5 attempts, the experiment started. Those who could not meet this requirement were excluded from data analysis (as a result, none of the participants were excluded based on this criterion). For data analysis, we excluded three participants due to experimental errors, and two participants because they deviated substantially from the prescribed tempo (outside 2 standard deviations from the average tempo across participants). Thirty-one participants (15 female) were included in data analysis. Most participants were right-handed (left: 2, ambidextrous: 2). They had 12.45 years of practice on average (\emph{SD} = 5.66). All participants gave their informed consent before the experiment started and received vouchers for their participation. The study was approved by the United Ethical Review Committee for Research in Psychology (EPKEB) in Hungary.

@@

We recruited 21 piano experts who already had a degree in (above bachelor or equivalent) in piano performance/teaching or were studying advanced piano performance at a music school at the time of recruitment. For data analysis, we excluded one participants due to insufficient motor skills, and one because of large tempo deviations (outside 2 standard deviations from the average tempo across participants). Nineteen participants (9 female) were included in the data analysis. Most participants were right-handed (left: 2). They had 15.65 years of practice on average (\emph{SD} = 5.67).

@@

The same apparatus as in Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contained our targeted expressions (i.e., articulation, dynamics) and was relatively simple in terms of motor skills. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in 4/4 meter. Non-expressive sheet music was used for the purpose of practice (\emph{Figure \ref{fig:stimuli-2}, A}). Expressive notations were added to the non-expressive sheet music for the experiment (\emph{Figure \ref{fig:stimuli-2}, B-C}). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as in Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., \emph{Figure \ref{fig:stimuli-2}, A}) prior to the experiment. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations (see details in \emph{\protect\hyperlink{supplemental}{Supplementary Materials}}). Third, participants could choose their preferred tempo from one of three options (100, 110 and 120 beats per minute). The chosen tempo was again cued by a leading metronome. In order to make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the non-expressive sheet music. As Experiment 1, those who could not perform the piece twice consecutively within 5 attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

Data cleaning, preprocessing and statistical analysis were almost identical to Experiment 1. For statistical analysis, only 8th notes with expressive notations were included. As a result, only one 8th note in the 4th measure without any expression was not included. IOIs were normalised by their preferred tempo because each participant chose a tempo from the three options. Given the different tempi, a key-overlap ratio (KOR) was calculated by dividing KOT by the mean IOI of each performance in order to normalise KOT. Additionally, we include KV difference (i.e., KV difference for each interval) at transition points (i.e., forte to piano or piano to forte) as a dependent variable based on the findings of Experiment 1. Three trials were entirely excluded from data analysis because participants did not follow the sheet music. Using the same approach as in Experiment 1, pitch errors were removed manually. For onsets, 3.32\% of the trials contained at least one pitch error (extra notes; 1.66\%, missing notes; 1.57\%, substituted notes; 0.09\%). For offsets, 5.11\% of the trials contained at least one pitch error (extra notes; 1.66\%, missing notes; 1.57\%, substituted notes; 1.88\%). For each dependent variable, removing outliers (i.e., responses outside 3 standard deviations from the mean) resulted in less than 5\% of overall responses being removed.
